Article Title

### 2014-05-17 plos.tex

Neural correlations do not come with ready or unambiguous mechanistic interpretations. They can arise from monosynaptic or polysynaptic interactions, common or correlated inputs, oscillations, top-down modulation, and background network fluctuations, and other mechanisms \cite{Perkel:1967, Moore:1970, Shadlen:1998, Salinas:2001, Ostojic:2009, Rosenbaum:2011}. But multineuronal recordings do provide more information than an equivalent number of separately recorded pairs of cells. For example, the eigenvalue decomposition of the covariance matrix expresses shared correlated activity components across the population; common fluctuations of population activity may be accurately represented by only a few eigenvectors that affect all correlation coefficients. On the other hand, a correlation matrix can be specified using the \emph{partial correlations} between pairs of the recorded neurons. The partial correlation coefficient between two neurons reflects their linear association conditioned on the activity of all the other recorded cells.  When a large fraction of the interacting variables are observed, partial correlations more closely reflect the direct causal effects between components of complex systems than correlations \cite{Whittaker:2009}. Partial correlations have been used to describe interactions between genes in functional genomics \cite{Schafer:2005, Peng:2009} and between brain regions in imaging studies \cite{Varoquaux:2012, Ryali:2012}. These opportunities have not yet been explored in neurophysiological studies, as most studies have only considered the distributions of pairwise correlations \cite{Zohary:1994, Bair:2001, Smith:2008, Ecker:2010}.

@@

Given an covariance matrix estimate, the correlation matrix $R$ is calculated by normalizing the rows and columns of $C$ by the square roots of its diagonal elements to produce unit entries on the diagonal:
\begin{equation}\label{eq:precision}
    R = \left(I\circ C\right)^{-\frac 1 2} C \left(I\circ C\right)^{-\frac 1 2},
\end{equation}
where $\circ$ denotes the entrywise matrix product (Hadamard product) and $I$ is the $p\times p$ identity matrix. 
Similarly, the matrix of partial correlations $P$ is computed by normalizing the \emph{precision matrix} $C^{-1}$ \cite{Whittaker:2009}:
\begin{equation}\label{eq:partial}
    P = -\left(I\circ C^{-1}\right)^{-\frac 1 2} C^{-1} \left(I\circ C^{-1}\right)^{-\frac 1 2}
>>>>>>> a0b18393448df9cb13967143cee42a00d25a7824
\end{equation}
Here $\circ$ denotes the entrywise matrix product (Hadamard product) and $I$ is the identity matrix.

\emph{Pairwise partial correlations} are less familiar in the neuroscience community. 
The partial correlation between a pair of variables is the Pearson correlation coefficient of the residuals of the linear least-squares predictor of their activity based on all the other variables, excluding the pair \cite{Anderson:2003, Whittaker:2009}. Partial correlations figure prominantly in probablistic \emph{graphical modelling} wherein the joint distribution is explain by sets of two-way interactions \cite{Whittaker:1990}. Under the multivariate Gaussian distribution, zero partial correlations are equivalent to conditional independence of the pair, implying a lack of direct interaction between them \cite{Dempster:1972, Whittaker:1990}. In the general case, partial correlations can serve as a measure of conditional independence under the assumption that most dependencies in the system include strong linear effects \cite{Whittaker:1999,Baba:2004}.

Pairwise partial correlations are closely linked to the \emph{precision matrix}, \ie the inverse of the covariance matrix \cite{Dempster:1972,Whittaker:1999}. Given the covariance estimate $C$, the matrix of partial correlations $P$ is computed by normalizing the rows and columns of the \emph{precision matrix} $C^{-1}$ to produce negative unit entries on the diagonal:
\begin{equation}\label{eq:partial}
    P = -\left(I\circ C^{-1}\right)^{-\frac 1 2} C^{-1} \left(I\circ C^{-1}\right)^{-\frac 1 2}
\end{equation}

@@

Estimator $C_{\sf factor}$ approximates the covariance matrix by the factor model $L + D$, where $L$ is the $p\times p$ positive \hl{semidefinite} matrix with low rank and $D$ is a diagonal matrix of independent variances. The estimator, 
\begin{equation}\label{eq:c-factor}
C_{\sf factor} = L + (1-\lambda)D + \lambda \bar D,
\end{equation}
has two hyperparameters: the rank of $L$ (\ie the number of latent factors) and the shrinkage intensity $\lambda$. The structure imposed by $C_{\sf factor}$ favors conditions in which the population activity is linearly driven by a number of latent factors that affect many cells while direct interactions between the recorded cells are insignificant (\figref{1}{Row 1, B}).

@@

Estimator $C_{\sf sparse+latent}$ has two hyperparameters: the number of latent units $d$ and \hl{hyperparameter $\lambda$ regulating the sparsity level}. $C_{\sf sparse+latent}$ estimates a larger sparse precision matrix $S^\ast$ of the joint distribution of the $p$ observed neurons and $d$ latent units.  
\begin{equation}
S^\ast=
\begin{pmatrix}
S & S_{12} \\
S_{12}^\T & S_{22}
\end{pmatrix},
\end{equation}
where the $p\times p$ partition $S$ corresponds to the visible units and expresses their partial correlation structure, and $S_{12}$ and $S_{22}$ are of size $p\times d$ and $d\times d$, respectively. 
Then the covariance matrix of the observed population is 
\begin{equation}
C_{\sf sparse+latent} = \left(S-S_{12}S_{22}^{-1}S_{12}^\T\right)^{-1}
\end{equation}
The $p\times p$  matrix $L=S_{12}S_{22}^{-1}S_{12}^\T$ has rank $d$. Rather than searching for the optimal sparse structure of $S_{12}$ and $S_{22}$, an ill-posed problem, we estimated these components together as the low-rank matrix $L$. We solved this problem by adapting the ADMM algorithm to minimizing the loss function with penalty on the $L_1$-norm  of $S$ and the nuclear norm of $L$ \cite{Chandrasekaran:2010,Ma:2013}:
\begin{equation}
(S,L) = \argmin\limits_{\hat S,\hat L} \loss{\hat S-\hat L, C_{\sf sample}} + \alpha\|\hat S\|_1 + \beta \Tr{\hat L}
\end{equation}

>>>>

Neural correlations do not come with ready or unambiguous mechanistic interpretations. They can arise from monosynaptic or polysynaptic interactions, common or correlated inputs, oscillations, top-down modulation, and background network fluctuations, and other mechanisms \cite{Perkel:1967, Moore:1970, Shadlen:1998, Salinas:2001, Ostojic:2009, Rosenbaum:2011}. But multineuronal recordings do provide more information than an equivalent number of separately recorded pairs of cells. For example, the eigenvalue decomposition of the covariance matrix expresses shared correlated activity components across the population; common fluctuations of population activity may be accurately represented by only a few eigenvectors that affect all correlation coefficients. On the other hand, a correlation matrix can be specified using the \emph{partial correlations} between pairs of the recorded neurons. The partial correlation coefficient between two neurons reflects their linear association conditioned on the activity of all the other recorded cells \cite{Whittaker:1990}.  Under some assumptions, partial correlations measure conditional independence between variables and may more directly approximate causal effects between components of complex systems than correlations \cite{Whittaker:1990}. For this reason, partial correlations have been used to describe interactions between genes in functional genomics \cite{Schafer:2005, Peng:2009} and between brain regions in imaging studies \cite{Varoquaux:2012, Ryali:2012}. These opportunities have not yet been explored in neurophysiological studies where most analyses have only considered the distributions of pairwise correlations \cite{Zohary:1994, Bair:2001, Smith:2008, Ecker:2010}.

@@

Given an covariance matrix estimate, the correlation matrix $R$ is calculated by normalizing the rows and columns of $C$ by the square roots of its diagonal elements to produce unit entries on the diagonal:
\begin{equation}\label{eq:precision}
    R = \left(I\circ C\right)^{-\frac 1 2} C \left(I\circ C\right)^{-\frac 1 2},
\end{equation}
where $\circ$ denotes the entrywise matrix product (Hadamard product) and $I$ is the $p\times p$ identity matrix.

\emph{Pairwise partial correlations} are less familiar in the neuroscience community. 
The partial correlation between a pair of variables is the Pearson correlation coefficient of the residuals of the linear least-squares predictor of their activity based on all the other variables, excluding the pair \cite{Anderson:2003, Whittaker:1990}. Partial correlations figure prominantly in probablistic \emph{graphical modelling} wherein the joint distribution is explain by sets of two-way interactions \cite{Whittaker:1990}. Under the multivariate Gaussian distribution, zero partial correlations are equivalent to conditional independence of the pair, implying a lack of direct interaction between them \cite{Dempster:1972, Whittaker:1990}. In the general case, partial correlations can serve as a measure of conditional independence under the assumption that most dependencies in the system include strong linear effects \cite{Whittaker:1990,Baba:2004}. As neural recordings become increasingly dense, partial correlations may prove useful measures of conditional independence and (lack of functinonal connectivity) between pairs of neurons.

Pairwise partial correlations are closely linked to the elements of the \emph{precision matrix}, \ie the inverse of the covariance matrix \cite{Dempster:1972,Whittaker:1990}. Zero elements in the precision matrix signify zero partial correlation between the two variables. Given the covariance estimate $C$, the matrix of partial correlations $P$ is computed by normalizing the rows and columns of the \emph{precision matrix} $C^{-1}$ to produce negative unit entries on the diagonal:
\begin{equation}\label{eq:partial}
    P = -\left(I\circ C^{-1}\right)^{-\frac 1 2} C^{-1} \left(I\circ C^{-1}\right)^{-\frac 1 2}
\end{equation}

As the size of recorded neuronal populations increases, without substantial increases in recording durations, the \emph{condition number} of the sample covariance matrix also increases, making the partial correlation estimates \emph{ill-conditioned}: small errors in covariance estimates translate into much larger errors in partial correlations \cite{Ledoit:2004,Schafer:2005}. With massively multineuronal recordings, partial correlations cannot be estimated without \emph{regularization.}

@@

Estimator $C_{\sf factor}$ approximates the covariance matrix by the factor model $L + D$, where $L$ is the $p\times p$ positive semidefinite matrix with low rank and $D$ is a diagonal matrix of independent variances. The estimator, 
\begin{equation}\label{eq:c-factor}
C_{\sf factor} = L + (1-\lambda)D + \lambda \bar D,
\end{equation}
has two hyperparameters: the rank of $L$ (\ie the number of latent factors) and the shrinkage intensity $\lambda$. The structure imposed by $C_{\sf factor}$ favors conditions in which the population activity is linearly driven by a number of latent factors that affect many cells while direct interactions between the recorded cells are insignificant (\figref{1}{Row 1, B}).

@@

Estimator $C_{\sf sparse+latent}$ has two hyperparameters: the number of latent units $d$ and hyperparameter $\lambda$ regulating the sparsity level. $C_{\sf sparse+latent}$ estimates a larger sparse precision matrix $S^\ast$ of the joint distribution of the $p$ observed neurons and $d$ latent units.  
\begin{equation}
S^\ast=
\begin{pmatrix}
S & S_{12} \\
S_{12}^\T & S_{22}
\end{pmatrix},
\end{equation}
where the $p\times p$ partition $S$ corresponds to the visible units and expresses their partial correlation structure, and $S_{12}$ and $S_{22}$ are of size $p\times d$ and $d\times d$, respectively. 
Then the covariance matrix of the observed population is 
\begin{equation}
C_{\sf sparse+latent} = \left(S-S_{12}S_{22}^{-1}S_{12}^\T\right)^{-1}
\end{equation}
The $p\times p$  matrix $L=S_{12}S_{22}^{-1}S_{12}^\T$ has rank $d$. Rather than searching for the optimal sparse structure of $S_{12}$ and $S_{22}$, an ill-posed problem, we estimated these components together as the low-rank matrix $L$. We solved this problem by adapting the ADMM algorithm to minimizing the loss function with penalty on the $L_1$-norm  of $S$ and the nuclear norm of $L$ \cite{Chandrasekaran:2010,Ma:2013}:
\begin{equation}
(S,L) = \argmin\limits_{\hat S,\hat L} \loss{\hat S-\hat L, C_{\sf sample}} + \alpha\|\hat S\|_1 + \beta \Tr{\hat L}
\end{equation}
